Pipeline systems and infrastructure for the efficient and open processing of Big neuroimaging Data

### 2018-09-28 paper.tex

        Workflows, also known as pipelines, describe the flow of data from 
        activity to another. They are typically represented by a 
        Directed Acyclic Graph (DAG). Workflow activites, also referred to as 
        workflows stages, encapsulate a tool. Tools may either take the form of 
        a command-line tool or a custom script. The engines determine which
        workflow activity can be staged by determining if the stage dependencies
        have been satisfied.

        In order for a workflow engine to be successful, it must be simple to 
        compose workflows using that framework. New researchers should be able 
        to acquire the skills to compose worfklows in that framework without
        investing a significant amount of time in understanding it. Reuse of
        existing tools is also common occurrence in neuroinformatics and 
        workflow engines need to accomodate this by undertaking a modular 
        approach and being easily extensible. With the sharing of data also 
        comes the sharing of workflows. For labs to share their workflows with
        the community at large, these workflows must be easily understood by the
        community.

@@

        \section{Workflow language}\label{lang}
            The choice of workflow language (e.g. GUI, scripting language,
            programming language) greatly
            affects how easily it will get picked up by the community. Complex
            languages may mean greater efficiency of the the engines
            , but also result in extensive time spent on learning and developing
            the pipelines. Moreover, some researchers may not be well-versed in
            programming and may require very easy-to-use engines.

            To address this issue, some workflow engines have taken the route
            of employing a graphical user interface (GUI). This makes workflow 
            composition accessible to all researchers, including those with
            little to no programming background. The caveat to employing a 
            workflow engine with a GUI interface for workflow composition is
            that users are greatly limited by the tasks they can perform. 
            Workflow engines that employ a worfklow composition GUI include
            SPM, Taverna, OpenMOLE, LONI and Galaxy. \note{double-check that 
            none have a programming API.Loni does..i think SPM does too}

            As more experienced programmers may feel limited by a GUI, workflow
            engines have decided to provide users with the ability to describe
            their workflows in more flexible formats. Even those with GUIs for 
            workflow composition, such as LONI and OpenMOLE, allow more 
            experienced users to script their workflow. Six different 
            programming
            languages are used for describing scientific workflows: 1) Python, 
            2) MATLAB and Octave, 3) Shell, 4) XML, 6) Common Workflow Language 
            (CWL) and 6) Domain Specific Language (DSL). Workflows engines that 
            employ usage of Python include: Nipype, FastR, Pydpiper, Pegasus
            and Toil.
            Python is an excellent choice as a programming language for 
            scientific workflows, particularly neuroinformatics, due to its 
            widespread use within the community. Other advantages of 
            programming in Python including a wide array of available libraries,
            including neuroimaging-specific ones such as Nilearn and Nibabel
            \note{cite}, and an lower learning curve compared to other 
            programming languages. Additionally, Python is platform independent
            meaning it can be installed on any system. In terms of BigData 
            frameworks that employ Python, Dask is exclusively written in Python
            , whereas Hadoop MapReduce and Spark both have interfaces to Python
            while being written in Java and Scala, respectively.

            Another widely-used high-level programming language in the 
            neuroinformatics community is MATLAB/Octave. Workflow engines which
            use this language include: PSOM, SPM, and Automatic analysis. A 
            downfall to imposing MATLAB as programming language is that the 
            it is proprietary, thus imposing a fee on 
            anyone wishes to program in that language. To offset this, PSOM 
            ensure compatibility with the GNU Octave programming language, thus 
            removing the requirement to purchase software for workflow 
            composition. Similarly to Python, MATLAB and Octave are 
            high-level platform-independent languages. MATLAB, in particular,
            also provides a wide array of libraries, however, the software must
            be purchased in order to access them.

            Scientific workflows are typically composed in UNIX-based 
            environments. These environments typically have Make installed by
            default eliminating the need for users to install it manually.
            Furthermore, all users who use Linux environments will typically 
            need to write some form of Shell script. This is the basis to use 
            workflow engines, such as Make, as scientific workflow engines. 
            As Make is exclusive to UNIX environments, it is not 
            platform-independent. However, Make has the flexibility of being
            easily extensible and can interface directly with any command-line
            tool. Shell scripting, nevertheless, is generally not as intuitive
            as programming languages like Python and MATLAB.

            Composing workflows in XML is another option provided by some 
            workflow engines such as Pegasus and LONI. However, the use of a 
            markup language removes some of the flexibility. Users wanting to 
            include 
            custom scripts as workflow activities must do so by creating a 
            command-line interface for their custom script. Pegasus, however,
            does provide many interfaces to describe the pipeline in including
            Java, Python and Perl.

            Another alternative is to compose workflows using Common Workflow
            Language (CWL). This can be seen in workflow engines like Toil. 
            CWL is not a programming language, but a specification for workflows
            . Files can either be written in JSON or YAML formats. CWL is 
            based on workflow engines like Make, but differs in that tasks
            are isolated and inputs and outputs must be explicit. Similarly to 
            make, CWL is only available on UNIX platforms. \note{cite cwl 
            website}

            Domain Specific Languages (DSL) are also used to compose workflows.
            DSLs are programming languages that have been adapted to specific
            domains. Workflow engine examples include OpenMOLE, which was 
            derived from Scala, and Netflow, which is derived from the Apache 
            Groovy programming language. While both are written on top of the 
            Java platform, Nextflow is only compatible with UNIX-based 
            environments. These programming languages are not otherwise commonly
            used in neuroimaging, meaning that users must learn specific 
            languages that would not otherwise be used in their field.

@@

            Additionally, BigData frameworks are better suited for the 
            processing of large images. Scientific workflows are generally 
            file-based with neuroimaging frameworks only providing parallelism
            on the file front. Sparks main abstraction, the Resilient 
            Distributed Dataset (RDD), permits users to process large datasets.
            A library based on the RDD, \textit{linealg}, allows users to create
            vectors and matrices and use common Python libraries, such as 
            Numpy and Scikit-learn to manipulate the data. Dask similarly 
            provides the Dask bag for file manipulation, and the Dask array for
            the manipulation of large matrices. Dask provides its own API for
            the processing of large matrices that is based on Numpy. 
            Both frameworks also provide data structures to manipulate tabular
            data: the Spark Dataframe and the Dask Dataframe. Unlike Spark, the
            Dask Dataframe is based off of the Pandas Dataframe API.

        \section{Modularity}\label{mod}
            Often times, users will want to integrate existing tools, in 
            addition to custom scripts, into their workflows. As workflow 
            engines, with the exception of Make, are typically not written in 
            Shell, it is necessary to create interfaces to these tools. These
            interfaces may range from simple calls to the command-line, to more 
            complex ones that track inputs, outputs and other parameters; ensure
            that all parameters are specified in the correct order and that the
            parameter and file types are all correct; capture all of the 
            standard input/output; and create folders to encapsulate task inputs
            and outputs. Workflow engines typically rely on more complex types 
            of interfaces in order to ensure correct functioning of the pipeline
            . As these interfaces can be rather complex, it is necessary for the 
            workflow engine to provide an intuitive way for users to create 
            interfaces for their custom tools.

            Early neuroimaging workflow engines, such as SPM, are monolithic, 
            and therefore, do not provide a mechanism for users to incorporate
            their custom tools. Workflow engines, such as AA, were thus 
            conceived to enable engines such as SPM to be more modular. AA 
            allows users to define their tasks through XML specification. AA's 
            XML tasklist is simply an ordered list of tasks which also includes
            details on analysis-specific parameters. Dependencies are
            inferred by the program through analysis of the module interfaces.
            Module interfaces consist of MATLAB user script and an XML 
            descriptor. The descriptor contains details on the input and output
            data stream types (e.g. epi, structural, dicom\_header) 
            and the desired domain (e.g. study, subject, session). During 
            execution, each task is given its own unique folder to operate on. 
            The engine will copy data to and extract data from this folder.

@@

            Pegasus is another workflow engine that uses an XML format to 
            define the CLI tools interface. The XML interfaces are typically 
            incorporated directly into the workflow DAX definition. However, a 
            separate multiline text-based 'Transformation Catalogue' file can
            be provided. Information encoded within the interfaces include
            tool identifier (typically namespace::version), the URL or filepath
            to the executable, site identifier of the executable's host, 
            executable type (e.g. installed or staged) and the profile. Pegasus
            can also execute containerized tools. For this, a container 
            identifier, image URL, site identifier of the image, and container
            profiles must also be defined \note{from website}.

@@

            Nipype's wrapper, written in Python, can encapsulate CLIs, SPM, 
            MATLAB and Python scripts. Attributes and methods that can be 
            defined for all interfaces include InputSpec, OutputSpec and a 
            Python dictionary of all expected outputs. InputSpec and OutputSpec
            are two Nipype classes. The former defines the input parameters of 
            a command and the second defines the outputs that are possibly 
            generated by the command. For CLIs, it is necessary to define the
            command. Additionally, Nipype will start X virtual framebuffer 
            before executing a CLI, if specified. For SPM interfaces, the job 
            type and name must be specified. Interfaces to MATLAB require 
            definition of the run function, while Python interfaces require
            the definition of the \textit{\_run\_interface} function. \note{from
            website}.

@@

            PSOM's tool interfaces are known as \textit{Bricks}. \textit{Bricks}
            can interface with any MATLAB function or CLI. Bricks require the
            definition of input and output files, in addition to a structure of
            program options. A useful feature found in PSOM is the ability 
            to perform a "dry-run" of the Brick without executing the command.

@@

>>>>

        Workflows, also known as pipelines, describe the flow of data from 
        activity to another. They are typically represented by a 
        Directed Acyclic Graph (DAG) \tristan{Not necessarily a DAG, see Nipype for instance, it has loops.
        I don't think you want to enter in a taxonomy of languages though, just say 'graph'}. Workflow activites, also referred to as 
        workflows stages, encapsulate a tool \tristan{Mention 'interface' here as an intro to sub-sec 2}. Tools may either take the form of 
        a command-line tool or a custom script \tristan{A script is a command-line tool. I think tools may also be web services (Taverna) or just 
        software libraries (Dask)}. The engines determine which
        workflow activity can be staged by determining if the stage dependencies
        have been satisfied.

        In order for a workflow engine to be successful, it must be simple to 
        compose workflows using that framework. New researchers should be able 
        to acquire the skills to compose worfklows in that framework without
        investing a significant amount of time in understanding it. Reuse of
        existing tools is also common occurrence in neuroinformatics and 
        workflow engines need to accomodate this by undertaking a modular 
        approach and being easily extensible. With the sharing of data also 
        comes the sharing of workflows. For labs to share their workflows with
        the community at large, these workflows must be easily understood by the
        community. \tristan{Perhaps mention that this also relates to the sharing
        of reproducible results}.

@@

        \section{Workflow language}\label{lang}
            The choice of workflow language (e.g. GUI, scripting language,
            programming language) greatly
            affects how easily it will get picked up by the community. Complex
            languages may mean greater efficiency of the engines
            , but also result in extensive time spent on learning and developing
            the pipelines. Moreover, some researchers may not be well-versed in
            programming and may require very easy-to-use engines.

            To address this issue, some workflow engines have taken the route
            of employing a graphical user interface (GUI). This makes workflow 
            composition accessible to all researchers, including those with
            little to no programming background. The caveat to employing a 
            workflow engine with a GUI interface for workflow composition is
            that users are greatly limited by the tasks they can perform. 
            Workflow engines that employ a worfklow composition GUI include
            SPM, Taverna, OpenMOLE, LONI and Galaxy. \note{double-check that 
            none have a programming API.Loni does..i think SPM does too}
            \tristan{My conclusion on GUIs is that they are suitable only
            for very simple pipelines. Perhaps the intro could give an idea
            of the types of pipelines encountered, showing a simple one and
            a more complex one. No way fmriprep is described in a GUI.}

            As more experienced programmers may feel limited by a GUI, workflow
            engines have decided to provide users with the ability to describe
            their workflows in more flexible formats. Even those with GUIs for 
            workflow composition, such as LONI and OpenMOLE, allow more 
            experienced users to script their workflow. Six different 
            programming
            languages are used for describing scientific workflows: 1) Python, 
            2) MATLAB and Octave \tristan{I think the official name is GNU/Octave}, 3) Shell \tristan{scripting languages}, 4) XML, 6) Common Workflow Language 
            (CWL) and 6) Domain Specific Language (DSL). \tristan{I don't see why CWL and XML don't qualify as DSL} Workflows engines that 
            employ usage of Python include: Nipype, FastR, Pydpiper, Pegasus
            and Toil \tristan{Dask? Spark?}.
            Python is an excellent choice as a programming language for 
            scientific workflows, particularly neuroinformatics, due to its 
            widespread use within the community. Other advantages of 
            programming in Python including a wide array of available libraries,
            including neuroimaging-specific ones such as Nilearn and Nibabel \tristan{mention scipy, sci-kit learn, etc}
            \note{cite}, and an lower learning curve compared to other 
            programming languages. Additionally, Python is platform independent
            meaning it can be installed on any system. In terms of BigData 
            frameworks that employ Python, Dask is exclusively written in Python
            , whereas Hadoop MapReduce and Spark both have interfaces to Python
            while being written in Java and Scala, respectively.

            Another widely-used high-level programming language in the 
            neuroinformatics community is MATLAB/Octave. Workflow engines which
            use this language include: PSOM, SPM, and Automatic analysis. A 
            downfall to imposing MATLAB as programming language is that the 
            it is proprietary, thus imposing a fee on 
            anyone wishes to program in that language \tristan{You should also mention the
            advantage of Matlab: easy to learn, integrated ecosystem, 
            performance, see 
            \url{https://www.mathworks.com/products/matlab/matlab-vs-python.html}}. 
            To offset this, PSOM ensure compatibility with the GNU 
            Octave programming language, thus 
            removing the requirement to purchase software for workflow 
            composition. Similarly to Python, MATLAB and Octave are 
            high-level platform-independent languages. MATLAB, in particular,
            also provides a wide array of libraries, however, the software must
            be purchased in order to access them.

            \tristan{This paragraph should start with ``Shell scripting ...". The structuration of this section by language
            is great, but the writing of this paragram doesn't put it forward.}Scientific workflows are typically composed in UNIX-based 
            environments. These environments typically have Make installed by
            default eliminating the need for users to install it manually.
            Furthermore, all users who use Linux environments will typically 
            need to write some form of Shell script. This is the basis to use 
            workflow engines, such as Make, as scientific workflow engines. 
            As Make is exclusive to UNIX environments, it is not 
            platform-independent. However, Make has the flexibility of being
            easily extensible and can interface directly with any command-line
            tool. Shell scripting, nevertheless, is generally not as intuitive
            as programming languages like Python and MATLAB.

            Composing workflows in XML is another option provided by some 
            workflow engines such as Pegasus and LONI. However, the use of a 
            markup language removes some of the flexibility. \tristan{Before you explain the downside, 
            you should say what are the advantages: mainly language-independent parsers and validation.} Users wanting to 
            include 
            custom scripts as workflow activities must do so by creating a 
            command-line interface for their custom script. Pegasus, however,
            does provide many interfaces to describe the pipeline in including
            Java, Python and Perl.

            Another alternative is to compose workflows using Common Workflow
            Language (CWL). This can be seen in workflow engines like Toil. 
            CWL is not a programming language, but a specification for workflows
            \tristan{It is a YAML-based DSL for workflows. Just like Pegasus' DAX is an XML-based DSL}. Files can either be written in JSON or YAML formats. CWL is 
            based on workflow engines like Make \tristan{? I don't think CWL relies on make, does it?}, but differs in that tasks
            are isolated and inputs and outputs must be explicit \tristan{inputs and outputs are somewhat explicit in a Makefile}. Similarly to 
            make, CWL is only available on UNIX platforms \tristan{strange, the CWL interpreter is in Python so I'd expect it to run anywhere}. \note{cite cwl 
            website}

            Domain Specific Languages (DSL) are also used to compose workflows.
            DSLs are programming languages that have been adapted to specific
            domains. Workflow engine examples include OpenMOLE, which was 
            derived from Scala, and Netflow \tristan{Nextflow?}, which is derived from the Apache 
            Groovy programming language. While both are written on top of the 
            Java platform, Nextflow is only compatible with UNIX-based \tristan{How about using Linux everywhere instead of UNIX-based?} 
            environments. These programming languages are not otherwise commonly
            used in neuroimaging, meaning that users must learn specific 
            languages that would not otherwise be used in their field.

@@

            \tristan{So this is now about data structures, not just language.
            This is really excellent, but this should perhaps go into a 
            specific sub-section. I think you could mention that 
            scientific workflow engines traditionally rely on files 
            only, and simple types, but that big data engines have 
            invested a lot in data structures, starting from key-value 
            pair representations in MR to RDD and Dask arrays. Spark is 
            actually built around RDDs, this is salient-enough to be 
            discussed a bit more. It actually makes a lot of sense that 
            data-focused engines are built around data structures 
            instead of control constructs. The ACM Com Spark paper makes it clear
            that unifying data structures is key to performance and interoperability, 
            because it allows data to remain in-memory instead of having to go back
            and forth to different environments. I would only mention linealg 
            in the description of Dask arrays. So, how about a 
            sub-section 'data structures' with 4 paragraphs: (1) 
            scientific and neuroimaging engines, (2) Big-Data, MR, (3) 
            Spark's RDDs and DataFrames, (4) Dask arrays, mentioning 
            linealg. } Additionally, BigData frameworks are better 
            suited for the 
            processing of large images. Scientific workflows are generally 
            file-based with neuroimaging frameworks only providing parallelism
            on the file front. Sparks main abstraction, the Resilient 
            Distributed Dataset (RDD), permits users to process large datasets.
            A library based on the RDD, \textit{linealg}, allows users to create
            vectors and matrices and use common Python libraries, such as 
            Numpy and Scikit-learn to manipulate the data. Dask similarly 
            provides the Dask bag for file manipulation, and the Dask array for
            the manipulation of large matrices. Dask provides its own API for
            the processing of large matrices that is based on Numpy. 
            Both frameworks also provide data structures to manipulate tabular
            data: the Spark Dataframe and the Dask Dataframe. Unlike Spark, the
            Dask Dataframe is based off of the Pandas Dataframe API.

        \section{Modularity}\label{mod}
            \tristan{Since this section is about interfaces, I would
            rather call it ``Interfaces" or ``Tool interfaces".}
            Often times, users will want to integrate existing tools, in 
            addition to custom scripts, into their workflows. As workflow 
            engines, with the exception of Make, are typically not written in 
            Shell, it is necessary to create interfaces to these tools. These
            interfaces may range from simple calls to the command-line, to more 
            complex ones that track inputs, outputs and other parameters; ensure
            that all parameters are specified in the correct order and that the
            parameter and file types are all correct; capture all of the 
            standard input/output; and create folders to encapsulate task inputs
            and outputs. Workflow engines typically rely on more complex types 
            of interfaces in order to ensure correct functioning of the pipeline
            . As these interfaces can be rather complex, it is necessary for the 
            workflow engine to provide an intuitive way for users to create 
            interfaces for their custom tools.

            Early neuroimaging workflow engines, such as SPM, are monolithic, 
            and therefore, do not provide a mechanism for users to incorporate
            their custom tools \tristan{Users can still integrate tools as Matlab functions
            doing system calls}. Workflow engines, such as AA, were thus 
            conceived to enable engines such as SPM to be more modular. AA 
            allows users to define their tasks through XML specification. AA's 
            XML tasklist is simply an ordered list of tasks which also includes
            details on analysis-specific parameters. Dependencies are
            inferred by the program through analysis of the module interfaces.
            Module interfaces consist of MATLAB user script and an XML 
            descriptor. The descriptor contains details on the input and output
            data stream types (e.g. epi, structural, dicom\_header) 
            and the desired domain (e.g. study, subject, session). During 
            execution, each task is given its own unique folder to operate on. 
            The engine will copy data to and extract data from this folder.

@@

            Pegasus is another workflow engine that uses an XML format to 
            define the CLI tools interface. The XML interfaces are typically 
            incorporated directly into the workflow DAX definition. However, a 
            separate multiline text-based 'Transformation Catalogue' file can
            be provided. Information encoded within the interfaces include
            tool identifier (typically namespace::version), the URL or filepath
            to the executable, site identifier of the executable's host, 
            executable type (e.g. installed or staged) and the profile. Pegasus
            can also execute containerized tools \tristan{how about the other ones?}. For this, a container 
            identifier, image URL, site identifier of the image, and container
            profiles must also be defined \note{from website}.

@@

            Nipype's wrapper, written in Python, can encapsulate CLIs, SPM, 
            MATLAB and Python scripts. Attributes and methods that can be 
            defined for all interfaces include InputSpec, OutputSpec and a 
            Python dictionary of all expected outputs. InputSpec and OutputSpec
            are two Nipype classes. The former defines the input parameters of 
            a command and the second defines the outputs that are possibly 
            generated by the command. For CLIs, it is necessary to define the
            command. Additionally, Nipype will start X virtual framebuffer 
            before executing a CLI, if specified \tristan{you need to explain why}. For SPM interfaces, the job 
            type and name must be specified. Interfaces to MATLAB require 
            definition of the run function, while Python interfaces require
            the definition of the \textit{\_run\_interface} function. \note{from
            website}.

@@

            PSOM's tool interfaces are known as \textit{Bricks}. \textit{Bricks}
            can interface with any MATLAB function or CLI. Bricks require the
            definition of input and output files, in addition to a structure of
            program options. A useful feature found in PSOM is the ability 
            to perform a ``dry-run" of the Brick without executing the command.

@@

            \tristan{Good material in here, but it lacks a bit structure.
            I like your structure of the previous sub-section, by language. I think
            here you could have two groups: (1) DSL descriptions (XML, JSON, etc) and (2)
            Python or Matlab descriptions. One advantage of DSLs is that they can 
            be more easily reused in different engines, while programming languages
            are more flexible, for instance Nipype allows outputs to be specified
            at runtime (dynamic number of inputs with names only known after
            the interface ran.)}
